---
title: "Subsetting data in R: a balance between safety and speed"
date: 2025-07-08
description: "Work in Progress"
image: ""
categories:
  - Decision modelling
  - R
# tags:
#   - Survival analysis
format: 
  html:
    toc-depth: 2
fig-cap-location: margin
reference-location: margin
# citation-location: margin
zotero: "phd-thesis"
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", fig.retina = 3,
                      fig.width = 6, fig.height = (6 * 0.618),
                      out.width = "90%", collapse = TRUE)
options(digits = 3, width = 300)
```

Within the week, I was testing a function for use in a combined microsimulation model of of three smoking-related cancers. My objective was to make the code run as safe as possible, such that I am able to detect when something isn't working well. One of the columns in the R data table of the model population (named `pop_model)`, was `smok_amt_yrs` . I had renamed this column from `smoke_amt` just hours ago. I needed to extract this vector from the data table for further calculations, so I mistakenly did

``` r
amount <- pop_model$smoke_amt
```

Previously, I had changed another column from `smoke_dur` to `smoke_yrs` , because I felt the suffix yrs better indicated the number of years that an individual had smoked. Again, I mistakenly did

``` r
duration <- pop_model[["smoke_dur"]]
```

The final output of the function was intended to be a column in pop_model that expresses the risk of onset of cancer based on the amount and number of years an individual had smoked. I got the following error

``` r
Error in [.data.table:
! RHS of assignment to existing column 'prob_inc_cancer' is zero length but not NULL. If you intend to delete the column use NULL....
```

On investigation, I realized that `pop_model$smoke_amt` returned the right vector, even though it was meant to be `pop_model$smoke_amt_yrs`; however, `pop_model[["smoke_dur"]]` silently returned `NULL`. When I changed the latter to `pop_model[, smoke_dur]` – it returned an error, but the function was notoriously slow. To understand what happened, we need to start from the basics, and the most basic of basics is this:

> There are three subsetting operators, `[[`, `[`, and `$`.

## Three subsetting operators - safety first

To explain this further, consider we have the following data table.

```{r}
if (!requireNamespace("data.table", quietly = TRUE)) {
  install.packages("data.table")
}

n <- 1e+06
set.seed(12345)

age <- sample(40:60, n, TRUE)
sex <- sample(c("Male", "Female"), n, TRUE, c(0.65, 0.35))
smok <- ifelse(
  sex == "Male",
  sample(x = c("Never", "Former", "Current"), size = n, replace = TRUE, prob = c(0.2, 0.5, 0.3)),
  sample(x = c("Never", "Former", "Current"), size = n, replace = TRUE, prob = c(0.3, 0.6, 0.1))
)

dt <- data.table::data.table(age, sex, smok)

head(dt)
```

Actually, i have decided that age_diag is a better descriptive term than just age. For illustration, assume that these changes are not made from the original source

```{r}
n <- 1e+06
set.seed(12345)

age_diag <- sample(40:60, n, TRUE)
sex <- sample(c("Male", "Female"), n, TRUE, c(0.65, 0.35))
smok <- ifelse(
  sex == "Male",
  sample(x = c("Never", "Former", "Current"), size = n, replace = TRUE, prob = c(0.2, 0.5, 0.3)),
  sample(x = c("Never", "Former", "Current"), size = n, replace = TRUE, prob = c(0.3, 0.6, 0.1))
)

dt <- data.table::data.table(age_diag, sex, smok)
```

Suppose I want to subset the age_diag column as a vector. And then I run the following code

```{r}
dt20 <- dt[1:20]
dt20$age
```

Even though I have changed the column name to age_diag, the above code will return the vector of ages. This is because the `$` operator does not throw an error when a column name is not found. Instead, R uses partial matching to find the closest match. In this case, it finds `age_diag` and returns the corresponding vector.Suppose I now decide to use the `[` operator to subset the age column. I run the following code

```{r, error=F}
#| error: true
dt20[, age]

```

This time, it returns an error, because the column name `age` does not exist in the data table. The `[` operator does not use partial matching, and it requires an exact match for the column name. Suppose I now decide to use the `[[` operator to subset the age column. I run the following code

```{r}
dt20[["age"]]

```

This time, it returns `NULL`, without throwing an error. This can lead to silent errors in your code, as you may not realize that you are working with a `NULL` value instead of the expected data. For instance, assume I want to create a new column in the data table that is five years after the age of diagnosis. I run the following code, and get the following result, but it is not what I expected. The code runs without error, but the result is of length zero, because `dt20[["age"]]` returned `NULL`. This is a silent error that can be difficult to detect.

```{r}
dt20[["age"]] + 5

```

What have we learned from this?

> The `[[` operator is the least safe – while it requires an exact match for the column name, but it can still return `NULL` without throwing an error. The `$` operator is the safer, because it uses partial matching and can lead to silent errors. The `[` operator is the safest, because it requires an exact match for the column name and throws an error if the column name does not exist.

## Three subsetting operators - need for speed

Now let's test the speed of the three subsetting operators. We will use the `microbenchmark` package to measure the time taken by each operator to subset the `smok` column from the `dt` data table with 1 million rows. `bracket2`, `dollar`, and `dt_j` formats respectively represent the `[` , `$` , and `[[` operators. We will run each operator 10000 times and measure the average time taken for each operation.

```{r}
if (!requireNamespace("microbenchmark", quietly = TRUE)) {
  install.packages("microbenchmark")
}
microbenchmark::microbenchmark(
  bracket2 = dt[["smok"]],
  dollar   = dt$smok,
  dt_j     = dt[, smok],
  times = 1000,
  unit = "milliseconds"
)

```

The results indicate that the `$` operator runs in 0.002255 milliseconds, the `[[` operator runs in 0.000451 ms, and the `[` operator runs in 0.378635 ms. The `$` operator is the fastest, followed by the `[[` operator, and then the `[` operator. The `$` operator is the fastest for a single named extraction, since it goes straight to the column with almost no overhead. The `[[` operator is a bit slower, since it goes through an extra dispatch step first[^1]. The `[` operator is by far the slowest on a `data.table`, since it has to parse the whole call (i.e., checking for row filters, grouping, and joins) before it even gets to pulling out the column.

[^1]: a "low-level extractor" typically means a function or operator that pulls a value out of a data structure with minimal extra layers of logic -- zero method dispatch, validation, or other operations that a "high-level" version would include.

This may not be a big deal for one-off operations, but in a microsimulation model with millions of operations, the difference in speed can add up to a significant amount of time. For instance, if you are running a microsimulation model with 1 million individuals and you need to extract columns from the data table 100 times throughout the model, the `$` operator would take approximately 0.3 milliseconds, while the `[` operator would take approximately 37.9 milliseconds. still not too bad. What if you ran a probabilistic sensitivity analysis (PSA) on that model with 10000 iterations, 5 sub-populations, comparing 4 interventions -- a total of $100*10000*5*4 = 20,000,000$ operations on the 1 million rows. The `$` operator would take approximately 45 seconds to complete these 20 million operations, while the `[` operator would take approximately 126 minutes.

Okay, 45 seconds is great, but subsetting is not the only operation in a microsimulation model. Can we do better? The answer is yes, we can do better. It turns out there is a low-level extractor[^2] that behaves like `[[` or `$`, but is much faster, called `.subset2`. Let's test the speed of the `.subset2` function against the other three subsetting operators.

[^2]: a "low-level extractor" typically means a function or operator that pulls a value out of a data structure with minimal extra layers of logic -- zero method dispatch, validation, or other operations that a "high-level" version would include.

```{r}
if (!requireNamespace("microbenchmark", quietly = TRUE)) {
  install.packages("microbenchmark")
}
microbenchmark::microbenchmark(
  bracket2 = dt[["smok"]],
  dollar   = dt$smok,
  dt_j     = dt[, smok],
  subset2  = .subset2(dt, "smok"),
  times = 1000,
  unit = "milliseconds"
)

```

Crazy! The `.subset2` function runs in 0.000082 milliseconds, which is 6 times faster than the `$` operator, 30 times faster than the `[[` operator, and 4634 times faster than the `[` operator. For a microsimulation model with 1 million individuals and 20 million operations, the `.subset2` function would take approximately 1.6 seconds to complete these 20 million operations, while the `$` operator would take approximately 9 seconds, the `[[` operator would take approximately45 seconds, and the `[` operator would take approximately 126 minutes! Thus `.subset2` function is the fastest of all the subsetting operators

In terms of safety, the `.subset2` function behaves like the `[[` operator, in that it requires an exact match for the column name and returns `NULL` if the column name does not exist. However, it is much faster than the `[[` operator, and it is also faster than the `$` operator. For example, back to the age_diag example, if I run the following code

```{r}
.subset2(dt20, "age")
```

It returns `NULL`, without throwing an error. This is the same behavior as the `[[` operator, and it can lead to silent errors in your code, as you may not realize that you are working with a `NULL` value instead of the expected data. So what's the best option for subsetting data in R? It depends on your priorities.

> If you prioritize safety over speed, and you just want to carry out a few operations, then the `[` operator is the best option. If you prioritize speed over safety, then the `.subset2` function is the best option.

What about readability? The `$` operator is the most readable, followed by the `[[` operator, and then the `[` operator. The `.subset2` function is the least readable; it is not as commonly used as the other three operators.

``` r
.subset2(dt20, "age")
dt20[, age]
dt20[["age"]]
dt20$age
```

However, if you are working with a large data table with multiple operations and you need to extract columns quickly, then the `.subset2` function is probably the best option.